---
title: "Sevare_Ragot_Bitco"
output: html_document
---
```{r}
library(readxl)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(scales)
```

```{r}
getwd()
setwd("C:/Users/Jadee/Desktop/EDA/FINPROJ/Eda_Project")
```

```{r}
data <- ("typhonn-impacts.xlsx")
```

# Total Families and People Affected per Typhoon
```{r}
data_affected_cleaned <- data_affected_cleaned %>%
  mutate(Families = as.numeric(gsub(",", "", Families)),
  Person = as.numeric(gsub(",", "", Person))
)
```

```{r}
total_families_person <- data_affected_cleaned %>%
group_by(Typhoon name) %>%
summarise(
  Total_Brgys = sum(Brgys., na.rm = TRUE),
  Total_Families = sum(Families, na.rm = TRUE),
  Total_Persons = sum(Person, na.rm = TRUE)
)

print = "total_families_person"

```{r}
total_families_long <- total_families_person %>%
pivot_longer(cols = c(Total_Families, Total_Persons),
names_to = "Type",
values_to = "Count")
```

```{r}
ggplot(total_families_long,
aes(x = Typhoon name, y = Count, fill = Type)) + geom_bar(stat = "identity") +
coord_flip() + scale_y_continuous(labels = comma) +
labs(title = "Families vs Persons Affected per Typhoon",
x = "Typhoon Name",
y = "Count") +

theme_minimal()
```

# Total Damaged Houses (Totally vs Partially) per Typhoon
```{r}
dataf_two_cleaned <- datadmg_two_cleaned %>%
mutate(
  Damaged_Houses_Totally = as.numeric(gsub(",", "", Damaged_Houses_Totally)),
  Damaged_Houses_Partially = as.numeric(gsub(",", "", Damaged_Houses_Partially))
)
```
```{r}
total_damaged_types <- dataf_two_cleaned %>%
group_by(Typhoon name) %>%
summarise(
Total_Totally_Damaged = sum(Damaged_Houses_Totally, na.rm = TRUE),
Total_Partially_Damaged = sum(Damaged_Houses_Partially, na.rm = TRUE)
)

print = "total_damaged_types"
```

```{r}
total_damaged_long <- total_damaged_types %>%
pivot_longer(cols = c(Total_Totally_Damaged, Total_Partially_Damaged),
names_to = "Type",
values_to = "Count")
```

```{r}
ggplot(total_damaged_long,
aes(x = Typhoon name, y = Count, fill = Type)) + geom_col(position = "dodge") +
coord_flip() + scale_y_continuous(labels = comma) +
labs(title = "Totally vs Partially Damaged Houses per Typhoon",
x = "Typhoon Name",
y = "Number of Houses") +

theme_minimal()
```

# Casualties Summary per Typhoon (Dead, Injured, Missing)
```{r}
dataf_two_cleaned <- dataf_two_cleaned %>%
mutate( 
  Casualties_Dead = as.numeric(gsub(",", "", Casualties_Dead)),
  Casualties_Injured = as.numeric(gsub(",", "", Casualties_Injured)),
  Casualties_Missing = as.numeric(gsub(",", "", Casualties_Missing))
)
```

```{r}
total_casualties <- dataf_two_cleaned %>%
group_by(Typhoon name) %>%
summarise(
  Dead = sum(Casualties_Dead, na.rm = TRUE),
  Injured = sum(Casualties_Injured, na.rm = TRUE),
  Missing = sum(Casualties_Missing, na.rm = TRUE)
)

print = "total_casualties"
```

```{r}
typhoon_example <- total_casualties_long %>% filter(Typhoon name == 
total_casualties_long$Typhoon name[1])
```

```{r}
ggplot(typhoon_example, aes(x = "", y = Count, fill = Type)) +
geom_col(width = 1) + coord_polar(theta = "y") +
labs(title = paste("Casualty Breakdown for", typhoon_example$Typhoon name[1])) +
theme_void() + scale_fill_manual(values = c("red", "orange", "yellow"))

theme_minimal()
```
