---
title: "DATA PREPARATION FOR FLOOD CONTROL PROJECTS DATASET"
output: html_document
---

```{r, message=FALSE}
# Load required libraries
library(tidyverse) 
library(ggplot2)
library(dplyr)
```

# I. Data Preparation

## 1. Data Loading and Initial Inspection

Load the raw flood control project dataset and understand its structure.

```{r}
# Load flood control project (fcp) dataset
fcp_data <- read.csv("flood-control-projects-table_2025-11-21.csv")
head(fcp_data)
```

```{r}
# Quick peek 
glimpse(fcp_data)
```

*Observation: There are 36 columns and 9,855 records; most of the variable types are correct.*

```{r}
# Check for missing values
colSums(is.na(fcp_data))
```

*Observation: Only 2 variables, `Program` and `CompletionDateOriginal` have missing values. `Program` has no records at all, while `CompletionDateOriginal` only has 4. This is unlikely to cause any issues, as these columns will not be included in the filtered dataset used for the exploration.*

### 1.1 Data Dictionary – Original Flood Control Project Dataset

This data dictionary describes each variable in the dataset, including its data type, definition, format, and any relevant notes.

1.  **Identification and Administrative Fields**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **GlobalID** | Character | System-generated unique identifier for each record. | Always unique. |
| **ProjectID** | Character | Identifier assigned to the main flood control project. | Shared across components. |
| **ProjectComponentID** | Character | Identifier for the specific component or subproject. | One project may have multiple components. |
| **ProjectDescription** | Character | Narrative description of the flood control project. | Text field. |
| **ProjectComponentDescription** | Character | Detailed description of the specific project component. | Text field. |
| **Program** | Logical | Indicates program classification. | Mostly NA. |

2.  **Geographic Information**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **Region** | Character | Administrative region of project location. | e.g., “Region IV-B”. |
| **Province** | Character | Province where the project is located. | e.g., “Palawan”. |
| **Municipality** | Character | Municipality/City of implementation. | Full names provided. |
| **LegislativeDistrict** | Character | Congressional district of the project area. | Text format. |
| **Longitude** | Double | Longitude coordinate of project site. | WGS84. |
| **Latitude** | Double | Latitude coordinate of project site. | WGS84. |
| **X_geo** | Character | Geospatial object stored by GIS system. | Serialized GIS object. |

3.  **Project Classification**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **TypeofWork** | Character | Type of construction work performed. | e.g., “Construction of Flood Mitigation Structure”. |
| **infra_type** | Character | Infrastructure classification. | e.g., “Flood Control Structures”. |
| **type** | Character | General category label. | e.g., “flood_control”. |
| **slug** | Character | URL-friendly identifier for project description. | Lowercase hyphenated. |

4.  **Implementation Offices**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **ImplementingOffice** | Character | DPWH office responsible for project implementation. | District/Regional. |
| **DistrictEngineeringOffice** | Character | Specific DPWH District Engineering Office. | Matches implementing office. |

5.  **Financial Information**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **ABC** | Double | Approved Budget for the Contract (in PHP). | Numeric version. |
| **ContractCost** | Double | Actual contract cost (in PHP). | Numeric version. |
| **ABC_String** | Character | ABC stored as text. | Often no commas/formatting. |
| **ContractCost_String** | Character | Contract cost stored as text. | With decimals. |

6.  **Project Timeline**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **InfraYear** | Integer | Infrastructure year of the project. | Often same as funding year. |
| **FundingYear** | Integer | Year when project was funded. | Typically 2023–2024. |
| **CompletionYear** | Integer | Year indicated for project completion. | Derived from planned completion date. |
| **StartDate** | Character | Actual start date (MM/DD/YYYY). | Text format. |
| **CompletionDateOriginal** | Double | Original planned completion date stored as UNIX timestamp. | Requires conversion. |
| **CompletionDateActual** | Character | Actual completion date (YYYY-MM-DD). | ISO string. |

7.  **Contractor and Administrative Metadata**

| Variable Name | Data Type | Description | Notes |
|----|----|----|----|
| **ContractID** | Character | Contract number associated with the project. | Unique per contract. |
| **Contractor** | Character | Name of contractor executing the project. | Text field. |
| **ObjectId** | Integer | System object identifier for internal record tracking. | Not unique dataset-wide. |
| **CreationDate** | Double | Timestamp of record creation. | UNIX timestamp. |
| **Creator** | Character | System account that created the record. | Usually “dpwh_view”. |
| **EditDate** | Double | Timestamp of most recent edit. | UNIX timestamp. |
| **Editor** | Character | System account that edited the record. | Usually “dpwh_view”. |

## 2. Select Relevant Columns

Retain only the columns needed for analysis and geospatial tasks.

```{r}
fcp_geo <- fcp_geo %>%
  select(
    # Identification
    ProjectID, ProjectDescription,

    # Project Classification
    TypeofWork, infra_type,

    # Geographic Information
    Region, Province, Municipality, LegislativeDistrict,
    Longitude, Latitude,

    # Financial Information
    ABC, ContractCost,

    # Project Timeline
    FundingYear, CompletionYear
  )
```

*Observation: Features reduced from 36 → 14, which makes the dataset is more focused.*

### **2.1 Column Relevance**

#### **A. Identification**

-   **ProjectID:** Unique identifier for each project (for tracking and joining data).

-   **ProjectDescription:** Brief description of the project (for context in visualizations).

#### **B. Project Classification**

-   **TypeofWork:** Type of construction (e.g., "Construction of Flood Mitigation Structure")—helps categorize projects.

-   **infra_type:** Infrastructure classification (e.g., "Flood Control Structures")—useful for filtering and analysis.

#### **C. Geographic Information**

-   **Region, Province, Municipality, LegislativeDistrict:** Administrative boundaries for spatial aggregation and analysis.

-   **Longitude, Latitude:** Coordinates for mapping and spatial analysis.

#### **D. Financial Information**

-   **ABC (Approved Budget for the Contract):** Budget allocated for each project (for budget analysis by region).

-   **ContractCost:** Actual cost of the project (for cost vs. budget comparisons).

**E. Project Timeline**

-   **FundingYear:** Year the project was funded (for temporal filtering and trend analysis).

-   **CompletionYear:** Year the project was completed (for tracking project timelines).

### **2.2 Verify Quality of Filtered Data**

```{r}
glimpse(fcp_geo)
```

*Observation: After selecting only the relevant fields, the number of features was reduced from 36 to 14, resulting in a more focused dataset without unnecessary variables. All retained columns have appropriate data types, with numeric coordinates, properly encoded financial and temporal values, and categorical fields stored as character strings.*

## **3. Filter for 2019–Latest by `FundingYear`**

Year range of the data is filtered by `FundingYear` in order to show **where and when projects were initiated**, reflecting policy priorities and resource allocation. We filter this date for consistency with the other dataset explored in this study.

```{r}
fcp_geo <- fcp_geo %>%
  filter(FundingYear >= 2019)
```

### **3.1 Verify**

```{r}
range(fcp_geo$FundingYear, na.rm = TRUE)
```

### **3.2 Check the Number of Projects After Filtering**

```{r}
nrow(fcp_geo) #filtered dataset
nrow(fcp_data) #original dataset
```

*Observation: After filtering the original `fcp_data` by `FundingYear` which only include 2019-2025. From the original **9855** number of observations it went down to **9848**, meaning only **7 records** had a `FundingYear` outside 2019–2025.*

## **4. Save Cleaned Data**

Export the filtered and cleaned dataset for further analysis.

```{r}
write.csv(fcp_geo, "fcp_geo_2019_2025.csv", row.names = FALSE)
```

### 4.1 Metadata of `fcp_geo`

| **Column Name** | **Type** | **Description** |
|----|----|----|
| **ProjectID** | Character | Unique identifier for each project. |
| **ProjectDescription** | Character | Short description of the project’s scope or activity. |
| **TypeofWork** | Character | Type of construction work executed (e.g., flood mitigation). |
| **infra_type** | Character | General infrastructure classification (e.g., Flood Control Structures). |
| **Region** | Character | Administrative region of the project location. |
| **Province** | Character | Province where the project is implemented. |
| **Municipality** | Character | Municipality or city where the project is located. |
| **LegislativeDistrict** | Character | Legislative district of the project area. |
| **Longitude** | Double | Longitude coordinate of the project site (for spatial mapping). |
| **Latitude** | Double | Latitude coordinate of the project site (for spatial mapping). |
| **ABC** | Double | Approved Budget for the Contract (project allocation). |
| **ContractCost** | Double | Actual awarded contract cost. |
| **FundingYear** | Integer | Year the project received funding. |
| **CompletionYear** | Integer | Year the project was completed. |

# Summary of Data Preparation

-   Dataset reduced from 36 → 14 columns and filtered for FundingYear ≥ 2019.
-   Retained columns are relevant, clean, and correctly typed.
-   Minimal data loss occurred (only 7 rows removed).
-   The resulting dataset is ready for geospatial EDA and further analysis.
